---
title: "gdmmTMB"
output: github_document
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/dhercar/gdmmTMB/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dhercar/gdmmTMB/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The package `gdmmTMB` provides functions for fitting and interrogating Generalised Dissimilarity Mixed Models (GDMM) using maximum likelihood estimation, Laplace approximation and (optionally) Bayesian Bootstrapping.

# Installation
```{r, eval = FALSE}
devtools::install_github('dhercar/gdmmTMB')
```

# Example
We start with two matrices: A community matrix (site x species) and a matrix of environmental covariates. 
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(gdmmTMB)

# Community matrix
sp <- gdm::southwest %>% select(site, species) %>% 
  group_by(site, species) %>%
  summarise(n = 1) %>%
  pivot_wider(id_cols = site, names_from = species,values_from = n, values_fill = 0 ) %>%
  arrange(site)
```


```{r, message=FALSE, warning=FALSE, echo = FALSE}
kableExtra::kable(head(sp))
```


```{r, message=FALSE, warning=FALSE}
env <- gdm::southwest %>% select(-c(species)) %>%
  distinct() %>% arrange(site)
```


```{r, message=FALSE, warning=FALSE, echo = FALSE}
kableExtra::kable(head(env))
```


### Fitting model 

In this example, we specify a binomial distribution and estimate model parameters with a Bayesian bootstrapping approach. Monotonic I-splines for dissimilarity gradients (`diss_formula`) are specified via `isp(x)` and `mono = TRUE`.

```{r}
m <- gdmm(Y = sp[,-1],
          X = env,
          diss_formula = ~ 
            isp(bio6)  +
            isp(bio15) + 
            isp(bio19),
          family = 'binomial',
          n_boot = 100, # should be at least 1000
          bboot = TRUE,
          mono = TRUE)
```


### Model summary

```{r}
summary(m)
```


### Plot dissimilarity gradients 
```{r, message=FALSE}
f_x <- diss_gradient(m, CI_quant = c(0.5, 0.95))
```


```{r}
library(ggplot2)

do.call(rbind, f_x) %>% ggplot(aes(fill = var)) +
  theme_bw() +
  geom_line(aes(x = x, y = f_x)) +
  geom_ribbon(aes(x = x, ymax = `CI 2.5%`, ymin = `CI 97.5%`), alpha = 0.1) +
  geom_ribbon(aes(x = x, ymax = `CI 25%`, ymin = `CI 75%`), alpha = 0.3) +
  facet_wrap(~var, scales = 'free_x', strip.position = "bottom") +
  ylab('f(x)') +
  theme(legend.position = '',
        strip.placement = 'outside', 
        strip.background = element_blank(),
        aspect.ratio = 1)
```

